Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-04 submission.tex
More on optimization

%Linguistic motivation, insight, utility...

\paragraph{Morphological Annotation}
The morphological annotation and PoS tag set of NDT follows the Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which in turn is largely based on the work
of \newcite{Faa:Lie:Van:97}. The tag set consists of 12 morphosyntactic PoS
tags, with 7 additional tags for punctuation and symbols.  The tag set is thus
rather coarse-grained, with broad categories such as \texttt{subst} (noun) and
\texttt{verb} (verb). The PoS tags are complemented by a large set of
morphological features, providing information about morphological properties such as
definiteness, number and tense. These features are used in our tag set
modifications, where the coarse PoS tag of relevant tokens are concatenated
with one or more of these features to include more linguistic information in
the tags.

@@

\begin{table} \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    %\caption{Syntactic annotation choices in NDT.}
    \caption{Central head-dependent annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

@@

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic
setting, i.e. in application to raw text, we want to evaluate the parser on
automatically assigned PoS tags. For the training of the parser,
however, we have two options: using either gold standard or
automatically assigned tags. In order to settle on a configuration, we
conducted experiments with gold standard and automatically assigned
tags to see how they differ with respect to performance. The results
of our experiments reveal that the combination of training and testing
on automatic tags is superior to training on gold standard tags and
testing on automatic tags.  Consequently, the parser was both trained
and tested on automatically assigned tags in our experiments.

@@

%Note that it is absolutely crucial that the morphological features in the
%treebank are removed when using automatic tags, as they are still gold
%standard. For instance, if a verb token is erroneously tagged as a noun, we
%could potentially have a noun token with verbal features such as tense, which
%markedly obfuscates the training and parsing. Another important factor is that
%we want to isolate the effect of PoS tags, necessitating the exclusion of
%morphological features.

@@

\subsection{Initial/Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). The result of this was a total of 368 tags, which is clearly very
fine-grained. The two initial tag sets, i.e., the original tag set comprising
19 tags and the full tag set comprising 368 tags, thus represent two extremes
in terms of granularity.  To evaluate these tag sets and investigate how the tag
set granularity would affect the performance of tagging and parsing, we trained
and evaluated TnT and Mate on the training and development data, respectively,
on the two tag sets. In Table \ref{inittagseteval}, we report the results of
these experiments. We see that the tagger accuracy drastically drops when going
from the original to the full tag set. The MFT baseline for the original tag
set is 94.14\%, while it drops by almost 9 percentage points to 85.15\% for the
full tag set. TnT reports an accuracy of 97.47\% on the original tag set, which
is reduced to 93.48\% for the full tag set. These results confirm our
hypothesis that the very high linguistic quality in the full, fine-grained tag
set comes at the expense of drops in tagger performance. However, the
additional linguistic information provided by the full tag set improves the
parser performance. With the original tag set, Mate reports an LAS of 87.01\%
and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%, respectively,
when using the full tag set. As we are looking for linguistically informed
distinctions that improve the syntactic parsing, these results are promising
and serve to indicate that additional morphological information assists the
syntactic parsers, motivating the optimization of the existing PoS tag set.

@@

\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine, neuter),
definiteness (indefinite or definite) and number (singular or plural). There
is agreement in gender, definiteness and number between nouns and their
modifiers (adjectives and determiners).

\paragraph{Verbs}
Verbs are inflected for tense (infinitive, present, preterite, past perfect)
in Norwegian and can additionally take on mood (imperative, indicative) and
voice (active or passive).

\paragraph{Adjectives}
Adjectives agree with the noun they modify in terms of gender, number and
definiteness in Norwegian.

\paragraph{Determiners}
Like adjectives, determiners in Norwegian agree with the noun they modify in
terms of gender, number and definiteness.

>>>>

\paragraph{Morphological Annotation}
The morphological annotation and PoS tag set of NDT follows the Oslo-Bergen
Tagger \cite{Hag:Joh:Nok:00,Sol:13}, which in turn is largely based on the work
of \newcite{Faa:Lie:Van:97}. The tag set consists of 12 morphosyntactic PoS
tags, with 7 additional tags for punctuation and symbols. The tag set is thus
rather coarse-grained, with broad categories such as \texttt{subst} (noun) and
\texttt{verb} (verb). The PoS tags are complemented by a large set of
morphological features, providing information about morphological properties such as
definiteness, number and tense. These features are used in our tag set
modifications, where the coarse PoS tag of relevant tokens is concatenated with
one or more of these features to include more linguistic information in the
tags.

@@

\begin{table}
    \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    \caption{Central head-dependent annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

@@

\paragraph{Tags \& Features}
As we seek to quantify the effects of PoS tagging in a realistic setting, i.e.,
in application to raw text, we want to evaluate the parser on automatically
assigned PoS tags. For the training of the parser, however, we have two
options: using either gold standard or automatically assigned tags. In order to
settle on a configuration, we conducted experiments with gold standard and
automatically assigned tags to see how they differ with respect to performance.
The results of our experiments reveal that the combination of training and
testing on automatic tags is superior to training on gold standard tags and
testing on automatic tags.  Consequently, the parser was both trained and
tested on automatically assigned tags in our experiments.

@@

@@

\subsection{Initial/Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). The result of this was a total of 368 tags, which is clearly very
fine-grained. The two initial tag sets, i.e., the original tag set comprising
19 tags and the full tag set comprising 368 tags, thus represent two extremes
in terms of granularity.  To evaluate these tag sets and investigate how the tag
set granularity would affect the performance of tagging and parsing, we trained
and evaluated TnT and Mate on the training and development data, respectively,
on the two tag sets. In Table \ref{inittagseteval}, we report the results of
these experiments. We see that the tagger accuracy drastically drops when going
from the original to the full tag set. TnT reports an accuracy of 97.47\% on
the original tag set, which is reduced to 93.48\% for the full tag set. These
results confirm our hypothesis that the very high linguistic quality in the
full, fine-grained tag set comes at the expense of drops in tagger performance.
However, the additional linguistic information provided by the full tag set
improves the parser performance. With the original tag set, Mate reports an LAS
of 87.01\% and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%,
respectively, when using the full tag set. As we are looking for linguistically
informed distinctions that improve the syntactic parsing, these results are
promising and serve to indicate that additional morphological information
assists the syntactic parsers, motivating the optimization of the existing PoS
tag set.

@@

The morphological properties of the various parts-of-speech is reflected in the
morphological features associated with the respective PoS tags. In addition to
morphological properties such as definiteness, tense and number, all classes
except for verbs can be distinguished "on" type, e.g., common nouns and proper
nouns, while pronouns can be reflexive, reciprocal, personal or interrogative.

\subsection{Experiments}
\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine, neuter),
definiteness (indefinite or definite) and number (singular or plural). There
is agreement in gender, definiteness and number between nouns and their
modifiers (adjectives and determiners). Additionally, NDT has a separate case
feature to distinguishing nouns in genitive case.

Experimenting with nouns, we found that all tag set modifications yielded large
increases in LAS and UAS. The most informative features are definiteness, which
leads to an increase in LAS by 1.26 percentage points, to 88.27\%, and type,
yielding an LAS of 88.07\%. Turning to combinations of features, we found that
the combination of type and case, as well as type and definiteness, were the
most promising, which led us to combine type, case and definiteness in a final
experiment, resulting in LAS of 88.81\% and UAS of 91.73\%, constituting large
increases from parsing with the original tag set, 1.80 percentage points
and 1.54 percentage points, respectively.

\paragraph{Verbs}
Verbs are inflected for tense (infinitive, present, preterite, past perfect)
in Norwegian and can additionally take on mood (imperative, indicative) and
voice (active or passive). Note that voice and mood have only a single value,
\texttt{pass} (passive) and \texttt{imp} (imperative), respectively. Verbs
which are not passive are implicitly active, and verbs which are not imperative
are indicative.

Introducing features in isolation, mood is the only feature leading to
increases in LAS, with a reported LAS of 87.04\%. Imperative clauses are
fundamentally different from indicative clauses, as they lack an overt subject.
Tense yields an LAS of 86.97\% while voice yields an LAS of 86.96\%. Combining
mood and tense resulted in an LAS of 87.12\% and UAS of 90.31\%.

In an additional experiment, we mapped the verb tenses (mood, in the case of
imperative) to finiteness.  All verbs have finiteness, hence this distinction
has broad coverage.  This mapping is syntactically grounded as finite verbs and
nonfinite verbs appear in completely different syntactic construction, and
proved to greatly improve the parsing, as we saw the overall largest parser
accuracy scores, with 87.30\% for LAS and 90.43\% for UAS, 0.29 and 0.24
percentage points higher than the baseline, respectively.  This coincides with
the observations seen for Swedish in \newcite{Ovr:08}, where finiteness was found
to be a very beneficial linguistic feature for parsing.

\paragraph{Adjectives}
Adjectives agree with the noun they modify in terms of gender, number and
definiteness in Norwegian. Adjectives are also inflected for degree, either
positive, comparative or superlative.

\paragraph{Determiners}
Like adjectives, determiners in Norwegian agree with the noun they modify in
terms of gender, number and definiteness. Possessive pronouns are
treated/analyzed as determiners in NDT.
